Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-02-10 papers/wdmerger_I/paper.tex
Rewrite of the free-fall section so that it is in dimensionless units.

For a single star simulation, the star is simply placed at the center
of the computational domain, which we take to be the origin. For a
binary star simulation, we take as parameters the mass of the two
white dwarfs and the initial orbital period $T$. Using Kepler's third
law and assuming a circular orbit, we can then work out the orbital
separation $a$:
\begin{equation}
  a = \left(\frac{GM T^2}{4\pi^2}\right)^{1/3}.
\end{equation}
Here $M = M_P + M_S$ is the total mass of the system, where $M_P$ is
the specified \textit{primary} mass and $M_S$ is the specified
\textit{secondary} mass. The primary WD will always be on the left
side of the computational domain for our simulations, and is more
massive than the secondary. This reflects the usual terminology in the
literature where the primary WD is the accretor and the secondary is
the donor. The center of mass is located at the center of the
computational domain, and the stars lie along the $x$ axis, so that
the primary's center of mass is located at $x = -(M_S / M)\, a$ and
the secondary's center of mass is located at $x = (M_P / M)\, a$.

@@

A simple test to verify the Poisson solver implemented by CASTRO is
the case of gravitational free fall. In this setup, two stars, each
individually in an equilibrium state, are placed on the computational
grid, separated by an initial distance $r_0$ along the $x$ axis with
zero initial velocity. We choose stars of masses $0.6$ and $0.8\,
M_\odot$, with the lower mass star on the left side (e.g. $x < 0$) of
the domain and the higher mass star on the right, such that their
center of mass coincides with the center of the
domain. Gravitationally, the stars may be treated as point masses
until the point of contact, so the equation of motion governing the
distance $r$ between their centers of mass is that of simple free
fall:
\begin{equation}
  \ddot{r}(t) = - \frac{GM}{r},
\end{equation}
where $G$ is the gravitational constant and $M$ is the total mass of
the system. This differential equation has a closed-form solution for
the evolution time as a function of separation:
\begin{equation}
  t(r) = \sqrt{\frac{r_0^3}{2GM}} \left[ \text{arccos}\left(\sqrt{\frac{r}{r_0}}\,\right) + \sqrt{\frac{r}{r_0} \left(1 - \frac{r}{r_0}\right)}\ \right]. \label{analyticalFreeFall}
\end{equation}
This result can be derived by recognizing that
\begin{equation}
  t(r) = \int_{r_0}^r \frac{dr}{v(r)}
\end{equation}
and inserting the velocity as a function of radial separation,
\begin{equation}
  v(r) = \sqrt{\frac{2GM}{r_0}\left(\frac{r_0}{r} - 1\right)}.
\end{equation}
We determine the initial separation to be consistent with the
simulation performed in Section \ref{sec:kepler}; that is, we select
an initial orbital period and use Kepler's third law to calculate the
radius of a circular orbit with that period. We select an initial
orbital period of $T = 100$ s for this simulation, so that the separation is 
$r_0 = 3.61 \times 10^{9}\ \text{cm}$.
We chose a relatively low resolution simulation to demonstrate the
capabilities of CASTRO even while using modest resources. The
computational grid is covered by a coarse grid of $48^3$ zones, with
two levels of refinement above the coarse grid. Each refined grid
carries an increase in resolution by a factor of 4 relative to the
coarser grid below it. CASTRO initially assigns $93\%$ of the domain
to be covered by the intermediate resolution grids, and $0.05\%$ of
the domain to be covered by the finest resolution grids.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{freefall/plot_freefall}
  \caption{Time evolution of two initially stationary white dwarfs,
    mutually attracted to each other by the gravitational force. The
    horizontal axis gives the separation of the white dwarfs, scaled
    to the initial separation, and the vertical axis gives the elapsed
    time of the simulation, scaled to the total elapsed time
    necessarily. The solid curve shows the analytical result,
    calculated from Newtonian mechanics, and the circles show the
    samples from the time evolution with CASTRO. The agreement between
    theory and simulation is excellent, even for a simulation with
    modest resolution.}
  \label{Fig:Free Fall}
\end{figure}

The analytical result in Equation \ref{analyticalFreeFall} determines
the total elapsed free-fall time,
\begin{equation}
  t_{\text{ff}} = \frac{\pi}{2} \sqrt{\frac{r_0^3}{2GM}} = \frac{T}{4\sqrt{2}}.
\end{equation}
The physical radius of each white dwarf is roughly $10\%$ of the
initial separation, so we consider the evolution terminated when the
radial separation reaches that value (at that point, the stars will no
longer be in free-fall due to physical contact). Since the elapsed
time goes roughly as the square root of the distance for small $r$, we
consider the evolution terminated when $t = 0.99\, t_{\text{ff}}$. The
results of our simulation are shown in Figure \ref{Fig:Free Fall}. The
positions of the two stars are determined by calculating the center of
mass of the right $(x > 0)$ and left $(x < 0)$ sides of the domain at 
the end of each time step, and treating the centers of mass as the 
respective location of the two stars.

>>>>

For a single star simulation, the star is simply placed at the center
of the computational domain, which we take to be the origin. For a
binary star simulation, we take as parameters the mass of the two
white dwarfs and the initial orbital period $T$. Using Kepler's third
law and assuming a circular orbit, we can then work out the orbital
separation $a$:
\begin{equation}
  a = \left(\frac{GM T^2}{4\pi^2}\right)^{1/3}.
\end{equation}
Here $M = M_P + M_S$ is the total mass of the system, where $M_P$ is
the specified \textit{primary} mass and $M_S$ is the specified
\textit{secondary} mass. The primary WD will always start on the left
side of the computational domain for our simulations, and is more
massive than the secondary. This reflects the usual terminology in the
literature where the primary WD is the accretor and the secondary is
the donor. The center of mass is located at the center of the
computational domain, and the stars lie along the $x$ axis, so that
the primary's center of mass is located at $x = -(M_S / M)\, a$ and
the secondary's center of mass is located at $x = (M_P / M)\, a$.

@@

A simple dynamical test to verify the gravitation physics in CASTRO is
the case of gravitational free fall. We place two stars on the grid 
in the manner of Section \ref{sec:initial_models}. The distance $a$ between 
them corresponds to a chosen orbital period $T$, consistent with the total
system mass $M$, but we disable the rotational source terms so that 
the stars start at rest in an inertial reference frame. 
Thus the stars will simply begin moving toward each other.
As long as the stars remain approximately spherical, the stars can be 
treated as point masses (this only seriously breaks down after the stars
have come into contact). In dimensionless units where $r \to r / a$ and 
$t \to 2\sqrt{2}\pi t / T$, the simple free fall equation of motion governing the
distance $r$ between their centers of mass takes the form:
\begin{equation}
  \ddot{r}(t) = - \frac{1}{2r^2}.
\end{equation}
It is possible to derive a closed-form solution for the evolution time
as a function of separation by starting with the integral formulation,
\begin{equation}
  t(r) = \int_{1}^{r} \frac{dr}{v(r)}.
\end{equation}
The velocity $v$ (in dimensionless units) can be found by noting that 
$\ddot{r} = v\, dv / dr$ and then separating and integrating the equation 
of motion. This yields 
\begin{equation}
  v(r) = \sqrt{\left(\frac{1}{r} - 1\right)}.
\end{equation}
For our problem $0 \leq r \leq 1$, so this is always valid. Integrating, we find
\begin{equation}
  t(r) = \text{arccos}\left(\sqrt{r}\right) + \sqrt{r \left(1 - r\right)}. \label{analyticalFreeFall}
\end{equation}
so that the point of contact would occur at $t = 1$. We actually stop the simulation
at $t = 0.9$, which is when the effects from the extended sizes of the stars
starts to become important. The results of our simulation for a $128^3$ zone 
uniform grid are shown in Figure \ref{Fig:Free Fall}. They show excellent agreement
between the theory and the simulation results, even for this very modest resolution.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{freefall/plot_freefall}
  \caption{Time evolution of two initially stationary white dwarfs,
    mutually attracted to each other by the gravitational force. The
    horizontal axis gives the separation of the white dwarfs, scaled
    to the initial separation, and the vertical axis gives the elapsed
    time of the simulation, scaled to the time it would take two point masses
    to collide. The solid curve shows the analytical result,
    calculated from Newtonian mechanics, and the circles show the
    samples from the time evolution with CASTRO.}
  \label{Fig:Free Fall}
\end{figure}
